The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-29 rational.tex
intro

@@

This highly intuitive and appealing idea was further explored and refined by
many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet despite of its appeal, the idea
that continuations of this

appeared as functional
representation of labels and jumps,
Filinski, in 1989, observed that they
are the categorical introduced the symmetric $\lambda$-calculus

@@

In a seemingly unrelated development, Girard in 1987 introduced linear
logic~\cite{Girard87tcs} which, among other contributions, exposed an
additive/multiplicative distinction in logical connectives and rules. In
particular, linear logic includes additive disjunctions $\oplus$ and
conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
the additive connectives to each other (the dual of $\oplus$ is $\with$ and
vice-versa) and the multiplicative connectives to each other (the dual of
$\otimes$ is $\parr$ and vice-versa).

@@

%%%%%%%%%%%%%%%%%%
\subsection{Negative Types}

@@

ADD MATH.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

As with additive duals, we represent {{eta*}} and {{eps*}] using
`U'-shaped connectors. In the general case the wires for {{1}} will be
dropped as it can freely introduced or eliminated using {{uniti}} and
{{unite}}.

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{b^^^}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{b^^^}} is
always the dual of the chosen one.

\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eta_times1.pdf}
}
\end{center}

@@

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{b^^^}}. These two values are free to move independantly
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

\paragraph*{Entanglement and Anti-Particles.}
A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
quantum mechanics. The {{eta*}} is a site of fission that creates an
information particle and its anti-particle. Both flow in the same
direction in time. The produce particles are entangles in that actions
on one (such as transformation by application of an isomorphism)
affects the other. Further the fission site does not create a particle
in any specific state, but in a superposition of all possible states
as dictated by its type. In each possible world that the particle
exists, it takes on a specific value inhibiting its type and
correspondingly its entangled anti-particle takes the dual of the
specific value.

This also leads to the exiciting point of view that entangled
particle/anti-particle pairs act as functions in a physical
first-class sense. We can ask the fuction to be applied to a value by
unifying the anti-particle with a particle in a well-known state and
examining is pair. We shall use this point of view in the development
our SAT solver (see Sec. \ref{sec:sat-solver}).

\paragraph*{Coherence.} 
As before, the requisite coherence condition on Compact Closed
Categories is applicable and is expressed below in terms of
{{eta*}}/{{eps*}} and the multiplicative monoid.

\paragraph*{Annihilation.}
As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
additive {{trace+}}, {{trace*}} gives us the ability to find the
fixpoint of the contraint (expressed with combinators) that is
traced. If the constraint has not satisfying values, there is not
fixpoint possible -- we use the term `annihilation' to describe the
corresponding undefined state of the program. Annihilation and a
non-termination both describe undefined program execution -- while
non-termination characterizes an iteration that fails to terminate,
annihilation characterizes a constraint that has no fixpoint.

\paragraph*{Zero Information.}
As with {{eta+}}, we argue that {{eta*}} constructs no new
information, instead it gives us the ability to explore as many worlds
as permitted by the type {{b}}. So situate ourselves in any specific
world, we have to supply additional information in teh form of
unifying the particle or anti-particle with a particle representing
known information. Hence only by supplying information to a an
entangled pair can we read information from it -- and always exactly
as much information as we supplied.

ADD MATH.

@@

%%%%%%%%%%%%%%%%%%%
\subsection{SAT solver}
\label{sec:sat-solver}

>>>>

In a recent paper~\cite{infeffects}, we argued that, by encompassing
irreversible physical primitives, conventional abstract models of computation
have also inadvertently included some \emph{implicit} computational effects
which we called \emph{information effects}. We then developed a pure
reversible model of computation that is obtained from the type isomorphisms
and categorical structures that underlie models of linear logic and quantum
computing and that treats information as a linear resource that can neither
be erased nor duplicated. In this paper, we show that our pure reversible
model unveils deeper and more elegant symmetries of computation than have
previously been reported. In particular, we expose two notions of duality of
computation: an additive duality and a multiplicative duality that give rise
to negative types and fractional types respectively. Although these types
have previously appeared in the literature (see Sec.~\ref{sec:related}), they
have typically appeared in the context of conventional languages with
information effects, which obscured their appeal and murked their properties.

To get a feeling for the main idea, consider the following algebraic
manipulation relating a natural number $a$ to itself:
\[\begin{array}{rclclclclcl}
a &=& a + 0 &=& a + ((-a) + a) &=& (a + (-a)) + a &=& 0 + a &=& a
\end{array}\]
Although seemingly pointless, this isomorphism has a non-trivial and
interesting computational interpretation (see Sec.~\ref{sub:neg}).

corresponds to a type isomorphism between the object $a$ and itself.

Furthermore, we illustrate their true appeal and expressiveness is brought
forth by viewing them in the context of an information-preserving
computational model.

In addition, we show how these types enrich our computational model, they
obey the same laws as the rational numbers. In other words, every algebraic
identity that holds for the rational numbers corresponds to a type
isomorphism in our model.

Specifically, 
isomorphisms enrich our computational model with have an interesting
computational interpretation

 in our
model. For example,

On the left we have a producer
that can supply a value of type $a$ and on the right we have a consumer that
demands a value of type $a$. For the purposes of the example, think of the
producer as a buyer who has \$20.00 and the consumer as a seller who expects
\$20.00 for the merchandise. In a conventional computational model, one might
realize this situation by simply writing the identity function: the buyer
hands the money to the seller to finish the transaction. The above sequence
of isomorphisms implements this identity function in a much more interesting
way, however. 
of the above series of isomorphisms is that money is ``created'' on the
consumer side together with a corresponding debt that travels ``backwards''
until it is paid for by the money at the supplier. In other words, this
computation models a credit card transaction. It shares the same benefits of
such transactions: allowing transactions to occur without necessarily
demanding that the resources be present at the time of the transaction.

On the producer side, the debt is paid for by the money computation with
the identity function. the producer and consumer must somehow share an
explicit dependency that allows the value. However in our model, the presence
of negative types allows the produced value to satisfy the demand without the
producer or consumer even knowing about each other. As is explained in detail
in

Indeed, in 
the above example, the computation on the consumer side creates

In particular, linear logic~\cite{Girard87tcs}, among other contributions,
exposed an additive/multiplicative distinction in logical connectives and
rules. In particular, linear logic includes additive disjunctions $\oplus$
and conjunctions $\with$ as well as multiplicative disjunctions $\parr$ and
conjuctions $\otimes$. Duality is also prominent in linear logic: it relates
the additive connectives to each other (the dual of $\oplus$ is $\with$ and
vice-versa) and the multiplicative connectives to each other (the dual of
$\otimes$ is $\parr$ and vice-versa).

We furthermore demonstrate that, in our model,
programming with these negative and fractional types, is a new ``revolution''
breaking dependencies...

Since Filinski, we've had the
idea that values and continuations are like mirror images. In a conventional
language, the negative (continuation) side is implicit and we introduce
information effects on the positive. Trying to recover the duality from this
distorted positive side has always been messy. Now it looks clean because we
have kept the positive side pure.

\newpage

@@

This highly intuitive and appealing idea was further explored and refined by
many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet, despite its appeal, the duality
between values and continuations

@@

In a seemingly unrelated development, Girard in 1987 introduced linear

@@

%%%%%%%%%%%%%%%%%%
\subsection{Negative Types}
\label{sub:neg}

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection{Multiplicative Duality in {{langRev}} }

@@

@@

\paragraph*{Unification.}

test test test
\begin{multicols}{2}
\begin{center}
\scalebox{1.5}{
  \includegraphics{diagrams/eta_times1.pdf}
}
\end{center}

@@

\paragraph*{Entanglement.}

\paragraph*{Coherence.}

\paragraph*{Annihilation.}

\paragraph*{Zero Information.}

@@

%%%%%%%%%%%%%%%%%%%
\subsection{SAT solver}
